---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This assignment involves reading in and performing simple exploratory 
analyses of 2 months of a single person's Fitbit data which 
consists of estimated counts of steps taken in each 5 minute interval 
during the 61 days (from October 1 to November 30 of 2012).  We look at 
both total daily activity (summed up step counts), and the average 
(across days) step counts for each 5 minute interval.  These can tell us 
something about the person's activity patterns during the 2 month period.

## Loading and preprocessing the data

First, we want to load the `dplyr` package so we can process the data 
more easily:

```{r results="hide", message=FALSE}
library(dplyr)
```

The first step to actually processing the data is to unzip the 
`activity.zip` file:

```{r}
unzip("activity.zip")
```

This extracts the `activity.csv` file. 
We then want to read this file into a data frame, and I choose to reorder
the columns so that the observation ID variables (`date` and `interval`) 
are the first columns and the measurement variable (`steps`) is the last 
variable.

```{r}
dfr <- read.csv("activity.csv")
dfr <- select(dfr, date, interval, steps)
head(dfr)
tail(dfr)
```

One peculiarity to note about the interval data is that the indices are 
non-uniformly defined.

```{r}
unique(dfr$interval)
```

Hours begin on indices 0, 100,.. 2300, but the interval indices within each 
hour tell what the starting minute is (from 0-55) within the hour.

Now there are a couple of data aggregations we want to perform for our 
future analyses.  First, we want to pool by date and add up the steps for 
each day.

```{r}
daySteps <- group_by(dfr, date) %>% summarize(totalSteps=sum(steps, na.rm=TRUE))
head(data.frame(daySteps), 61)
```

We should note that the days with missing data end up having step counts 
of 0.

Next we want to pool by the 5 minute interval index and average across all 
days, to give average step counts for each interval.

```{r}
intervalAveSteps <- group_by(dfr, interval) %>% summarize(aveSteps=mean(steps, na.rm=TRUE))
head(data.frame(intervalAveSteps))
tail(data.frame(intervalAveSteps))
```

## What is mean total number of steps taken per day?

We want to start looking at statistics for the daily step counts and make 
a histogram of these to give us an idea how daily activity is distributed 
among the days.

```{r}
totStepsMean <- mean(daySteps$totalSteps)
totStepsMed <- median(daySteps$totalSteps)
hist(daySteps$totalSteps, main="Histogram of step counts for the 61 days", 
     xlab="Steps Taken During the Day", ylab="Frequency (days)")
```

The mean calculated for the daily sums is `r totStepsMean` and the median 
is `r totStepsMed`.  This is skewed by a lot of data points with 
low step counts, which happen to include the days where data was missing.

## What is the average daily activity pattern?

We want to get an idea of the daily trend of activity averaged across all 
days.  The first two lines below find the 5 minute interval index for the 
interval where the average over days is maximum.  The plot, then, shows the 
average step counts vs. interval index.

```{r}
rowInd <- which(intervalAveSteps$aveSteps==max(intervalAveSteps$aveSteps))
maxStepInterval <- intervalAveSteps$interval[rowInd]
plot(intervalAveSteps$interval, intervalAveSteps$aveSteps, type="l", 
     main="Average Activity vs. 5-Minute Interval", 
     xlab="5 Minute Interval Index", ylab="Average Activity (Steps)")
```

The interval index at which the average count is maximized is 
`r maxStepInterval` (7:35-7:40 am). Activity appears to pick up after 
interval 500, 
which would correspond to 4 am, which suggests that this peak 
corresponds to the person waking up and starting their day.

## Imputing missing values

The previous analyses have treated the missing data, effectively, as if that 
data had 0 steps taken in the intervals.  This led to there being a number 
of days with a sum of 0 steps taken, which is entirely unrealistic and will 
skew the data.  So, we may want to fill in this data with some *imputed* 
values which are good "guesses" as to typical values that might have been 
seen had valid data been collected. 

```{r, echo=FALSE}
numNAs <- sum(is.na(dfr$steps))
```

There are `r numNAs` data points (corresponding to 5 min intervals) where 
the data is missing.  Observation reveals that these are for 8 particular 
days, when data is missing.  (Perhaps the wearer did not wear their Fitbit 
during those days.)

I chose to use the 5 minute interval averages (across days) that the 
previous graph shows to serve as those default values during particular 
missing values.  In order to perform this imputation process, we create 
a new dataframe that merges the original data with the average counts. 
We then round these averages to the nearest integer, since step counts 
should take on integer values.  Then, for all of the NA step values, we 
copy over these new rounded values.  Finally, we sort the frame by date, 
then interval, and we pull out the columns we want.

```{r}
dfr.imp <- merge(dfr, intervalAveSteps, by.x="interval", by.y="interval")
dfr.imp <- mutate(dfr.imp, aveSteps=round(aveSteps))
dfr.imp[is.na(dfr.imp$steps), ]$steps <- dfr.imp[is.na(dfr.imp$steps), ]$aveSteps
dfr.imp <- arrange(dfr.imp, date, interval) %>% select(date, interval, steps)
head(dfr.imp)
tail(dfr.imp)
```

As can be seen, the step values at the start and end of the table, that were 
originally NA, have now been filled in.

Now, as before, we can take this data and calculate step sums for each day.

```{r}
daySteps.imp <- group_by(dfr.imp, date) %>% summarize(totalSteps=sum(steps))
head(data.frame(daySteps.imp), 61)
```

Now we calculate a new mean and median and plot a new histogram with this 
data.

```{r}
totStepsMean.imp <- mean(daySteps.imp$totalSteps)
totStepsMed.imp <- median(daySteps.imp$totalSteps)
hist(daySteps.imp$totalSteps, 
     main="Histogram of step counts for the 61 days (imputed data)", 
     xlab="Steps Taken During the Day", ylab="Frequency (days)")
```

The new mean calculated for the daily sums with the imputed data is 
`r totStepsMean.imp` and the median is `r totStepsMed.imp`. The mean has 
significantly increased and is now much closer to the median, which 
indicates an "unskewing" of the distribution. 
One of the first things to notice here is that 
the histogram is far more symmetric, i.e., less skewed, than the previous 
histogram that used the non-imputed missing data.  This illustrates the 
benefit of missing data imputation: you arrive at more realistic values 
that lead to more representative distributions.

## Are there differences in activity patterns between weekdays and weekends?

In the previous line plot, we viewed the average step count vs. interval 
index, with the data averaged over all days.  But what if we want to 
compare the averages on the weekday vs. the weekend?

First we calculate a new weekend variable to add to our (imputed) data. 
Then, we perform a new aggregation by groups that include both interval 
index and "weektime" (weekday vs. weekend), and we calculate averages for 
each of the groups.

```{r}
isWeekend <- weekdays(as.Date(as.character(dfr.imp$date))) %in% c("Saturday", "Sunday")
dfr.imp <- mutate(dfr.imp, weektime=factor(sapply(isWeekend, 
    function(x) {if (x) "weekend" else "weekday"})))
intervalWTAveSteps <- group_by(dfr.imp, interval, weektime) %>%
    summarize(aveSteps=mean(steps))
head(data.frame(intervalWTAveSteps))
tail(data.frame(intervalWTAveSteps))
```

This data is organized in a way that is conducive to using ggplot to 
make a panel plot.  We plot two panels (weekday vs. weekend) and, for each 
of these, the average step count vs. the interval index.

```{r}
library(ggplot2)
qplot(interval, aveSteps, data=intervalWTAveSteps, facets=. ~ weektime, 
      geom=c("point", "line")) + 
    labs(title="Average Activity (Weekday and Weekend) vs. 5-Minute Interval") +
    labs(x="5 Minute Interval Index") + labs(y="Average Activity (Steps)")
```

Overall, there appears to be more (average) activity during certain parts 
of the day during weekdays than on the weekends (for example between 
intervals 500 and 1000). Since this corresponds to the time between 4 and 
9 am, one likely hypothesis is that the wearer is sleeping in late during 
the weekends.